---
title: "Initial GIS processing"
---
 

```{r, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
knitr::opts_knit$set(root.dir = "./eden_data")
```


## Aim

To explore and pre-process the GIS data for the Eden catchment.

## Required R Packages

R has multiple packages for the processing and analysis spatial data. A good
overview is given in the [Spatial Task View on
CRAN](https://cran.r-project.org/web/views/Spatial.html).
For this part of the training course we use a set of mature package
`terra` which is a dependancy of `dynatopGIS`. Similar results could be genratered
with other packages.

Attached the `terra` package to the R environment so that its
functions are available.
```{r attach_gis}
library(terra)
library(dynatopGIS)
```

and remember to move to the eden_data directory e.g.

```{r, eval=FALSE}
setwd("../eden_data")
```

## Data Required by `dynatopGIS`

To build a Dynamic TOPMODEL the `dynatopGIS` package requires the following

- A raster map of the catchment, possiblly deliniating subcatchments
- A raster DEM covering the catchment
- A vector representation of the river network

Optionally the user can also provide

- A raster map deliniating spatial rainfall patterns
- A raster map deliniating spatial PET patterns
- Raster maps containing further data to help classify the HRUs

All raster data must in the same resolution and projection as the catchment map. The vector river data must have the same projection.

> **WARNING**: `dynatopGIS` only works with
> projected raster data with square cells!

> **WARNING**: `dyntopGIS` presumes all raster values are numeric!

> **Note** The raster map of the catchment should have only `NA` values in the
> edge row and columns.

## Loading Data

There are four types of spatial data in the example; raster, polygons, lines and
points; in three different formats; Geotiff, shapefile and csv.

To start use functions in the raster package to load both the raster
data and that contained in the shapefiles:

```{r, read_raster}
dem <- rast(file.path(".","unprocessed","Eden_DEM.tif")) # load the dem as a raster layer
eden <- vect(file.path(".","unprocessed","Eden_Catchment.shp")) # load the outline of the sub-catchments from the shapefile
channel <- vect(file.path(".","unprocessed","Eden_River_Network.shp")) # load the river network from the shapefile
urban <- vect(file.path(".","unprocessed","Eden_Urban.shp")) # load the urban area map from the shapefile
```

Typing the variable name at the R prompt will display a summary of the data
loaded, for example

```{r, print_raster}
dem
```

Compare the `dem`, `eden`, `channel` and `urban` variable to check that:

- there is a common projection
- the cells of DEM are square

Reading the gauge locations from the csv file is a two stage process

```{r, read_csv}
 # read in the csv file to a date.frame
csvGauges <- read.csv(file.path(".","unprocessed","Eden_Gauge_Sites.csv"))
# convert the data frame to a SpatVect object - take projection from the DEM
gauges <- vect(csvGauges,geom=c("x","y"),crs=crs(dem))
```

Next plot the data using the default methods in the raster package

```{r data_plot}
plot(dem) # underlying coloured image with scale
plot(urban,add=TRUE,col="grey") # grey polygons
plot(channel,add=TRUE,col="blue") # channels as blue lines
plot(eden,add=TRUE) # outlines of sub-catchments in black
plot(gauges,add=TRUE,col="red") # gauges as red +
```

## Rasterising Layers

The basis of the landscape discretisation is the raster catchment map. In this
example, as in many situation the catchment outlines are vectors shapes. These
are rasterised using the grid provided by the raster DEM. In the catment map each
subcatchment requires a unique numeric value which is provided by the `id`
feild.

```{r rasterise_subcatch}
## show data frame with id values
head(eden)
## rasterise
eden_raster <- rasterize(eden, dem, field="id")
## plot the resulting raster
plot(eden_raster)
```

To ensure the edges are `NA` values the raster is extended, along with that of
the DEM

```{r extend_rast}
eden_raster <- extend(eden_raster,1)
dem <- extend(dem,1)
```

Since the urban areas may be used in the classification
they are also rasterised. Suitable numeric values are taken from the unique identifiers in the
data.

In the case of the urban data there are two possible identifiers objectid and bua_id:
```{r raterise_urban}
## show data frame with id values
head(urban)
## rasterise using objectid as the raster value
urban_raster <- rasterize(urban, eden_raster, field="objectid")
## plot the resulting raster
plot(urban_raster)
```

The raster fields created (`eden_raster`, `urban_raster` and extended `dem`) exist only in memory (or
temporary files). Since they will be needed later we will save them:

```{r, save_eden_raster}
writeRaster(eden_raster,file.path(".","processed","eden.tif"),overwrite=TRUE)
writeRaster(urban_raster,file.path(".","processed","urban.tif"),overwrite=TRUE)
writeRaster(dem,file.path(".","processed","dem.tif"),overwrite=TRUE)
```

## Channel Network

The representation of the channel network in dynatop is as a series of
connected reach lengths. In the processing done by dynatopGIS Each reach is
represented by a single spatial polygon. The data for each reach must include
the following feilds: 

 - name - The unique identifier/name of the reach
 - length - Length of the reach [m]
 - startNode - An identifier for the head of the reach
 - endNode - An identifier for the foot of the reach
 - width - Channel width [m]
 - slope - Channel slope [m/m]
 
 The startNode and endNode values are used to define the channel connectivity,
 while the physical properties are used in the `dynatop` numerical solutions.
 
 Looking at the channel data

```{r channel_data}
channel
```

we see that it already contains

- an identifier for the foot of the reach called endNode
- an identifier for the head of the reach called startNode
- the reach length
- a unique identifier called identifier

but is composed to spatial lines rather then polygons.

The `dynatopGIS` package provides a function for convienitently converting
line obejcts to the correct format.

```{r convert_channel}
conv_channel <- convert_channel(
    channel,
    property_names = c(name = "identifier", length = "length", startNode = "startNode",
                       endNode= "endNode"),
    default_width = 2,
    default_slope = 0.001
)
```

Thw two warning messages reveal that, since they are not specified, default
width and slope values are added. The width propoerty is used to buffer the
channel resulting in spatial poygons.

```{r conv_channel}
conv_channel
```

The converted channel data is saved for later use

```{r, save_eden_channel}
writeVector(conv_channel,file.path(".","processed","channel.shp"),overwrite=TRUE)
```

## Locating Gauges on the Channel Network

Before locating the gauges on the network note that in the map of the loaded data there is a single gauge outside to the catchment
area. As simple example of manipulating spatial data in R is to remove this by
cropping the gauges to the catchment area. This is done by

```{r crop_gauges}
gauges <- crop(gauges,eden)
```

To determine the location of the gauges on the river network we identify for each gauge the
closest reach. The identifier of this reach, along with the distance to
the gauge data is then added to the gauge data.

```{r locate_gauges}
dst <- distance(gauges,conv_channel) ## distance between gauges(row) and channels(column)
idx <- apply(dst,1,which.min) ## index of closest channel for each gauge

gauges$chn_identifier <- conv_channel$name[ apply(dst,1,which.min) ]
gauges$chn_distance <- apply(dst,1,min) ## distance to the channel
 
```

To check the results by plotting:

```{r crop_gauges_plot}
selected <- conv_channel[conv_channel$name %in% gauges$chn_identifier,]
plot(eden) # outlines of subcatchments in black
plot(conv_channel,add=TRUE,col="blue",border="blue") # channel network in blue
plot(selected,add=TRUE,col="orange",border="orange") # selected reachs in orange
plot(gauges,add=TRUE,col="red",pch=21) # gauges as red filled circles
```

The cropped list of gauges with channel identifier can be saved to a shapefile:

```{r save_cropped_gauges}
writeVector(gauges,file.path(".","processed","gauges.shp"),overwrite=TRUE)
```

> The quality of results of any method used for any automatic method of
> locating the gauges depends upon the accuracy of the GIS data. 

> In the
> numeric solution used in dyantop it is assumed that a gauge is at the foot of
> a reach. This should be allowed from when conceptualising the channel reaches.
