---
title: "Initial GIS processing"
---
 

```{r, include = FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
knitr::opts_knit$set(root.dir = "./eden_data")
```


## Aim

To explore and pre-process the GIS data for the Eden catchment.

## Installing required R packages

R has multiple packages for the processing and analysis spatial data. A good
overview is given in the [Spatial Task View on
CRAN](https://cran.r-project.org/web/views/Spatial.html).
For this
part of the training course we use a set of mature packages, rgeos, rgdal, and
raster. Similar results could be obtained using either of the terra or sf
packages.

To install the packages enter at the R command prompt
```{r install_gis, eval=FALSE}
install.packages(c("rgeos","rgdal","raster","sp","ncdf4"))
```
Once installed the raster and sp packages are attached to the R environment so that there
functions are available.
```{r attach_gis}
library(raster)
library(sp)
```

## Loading Data

There are four types of spatial data in the example; raster, polygons, lines and
points; in three different formats; Geotiff, shapefile and csv.

To start use functions in the raster package to load both the raster
data and that contained in the shapefiles:

```{r, read_raster}
dem <- raster(file.path(".","unprocessed","Eden_DEM.tif")) # load the dem as a raster layer
eden <- shapefile(file.path(".","unprocessed","Eden_Catchment")) # load the outline of the sub-catchments from the shapefile
channel <- shapefile(file.path(".","unprocessed","Eden_River_Network")) # load the river network from the shapefile
urban <- shapefile(file.path(".","unprocessed","Eden_Urban")) # load the urban area map from the shapefile
```

Typing the variable name at the R prompt will display a summary of the data
loaded, for example

```{r, print_raster}
dem
```

Comparing between the `dem`, `eden`, `channel` and `urban` variable shows
that:

- there is a common projection
- the cells of DEM are square

> **WARNING**: dynatopGIS only works with
> projected raster data with square cells! It is best to make sure you DEM has
> these properties before starting any analysis.

Reading the gauge locations from the csv file is a two stage process

```{r, read_csv}
 # read in the csv file to a date.frame
csvGauges <- read.csv(file.path(".","unprocessed","Eden_Gauge_Sites.csv"))
# convert the data frame to a SpatialPointsDataFrame - take projection from the DEM
gauges <- SpatialPointsDataFrame(coords=csvGauges[,c("x","y")],proj4string=crs(dem),data=csvGauges)
```

Next plot the data using the default methods in the raster package

```{r data_plot}
plot(dem) # underlying coloured image with scale
plot(urban,add=TRUE,col="grey") # grey polygons
plot(channel,add=TRUE,col="blue") # channels as blue lines
plot(eden,add=TRUE) # outlines of sub-catchments in black
plot(gauges,add=TRUE,col="red") # gauges as red +
```

## Rasterising Layers

The basis of the landscape discretisation is the DEM of the catchment. Any
information we may want to use in the classification of the catchment leading to the definition
of the Hydrological Response Units has to be provided as a raster of the
same resolution and projection as the DEM.

Since both the urban areas and sub-catchments may be used in the classification
both are rasterised. In R all raster values must be numeric. For both the urban
and sub-catchments the values are taken from the unique identifiers in the
data.

In the case of the urban data there are two possible identifiers objectid and bua_id:
```{r raterise_urban}
## show data frame with id values
head(urban@data)
## rasterise using objectid as the raster value
urban_raster <- rasterize(urban, dem, field="objectid")
## plot the resulting raster
plot(urban_raster)
```

For the sub-catchments the values are set to the numeric id:

```{r rasterise_subcatch}
## show data frame with id values
head(eden@data)
## rasterise
eden_raster <- rasterize(eden, dem, field="id")
## plot the resulting raster
plot(eden_raster)
```

The raster fields created (`eden_raster`, `urban_raster`) exist only in memory (or
temporary files). Since they will be needed later we will save them:

```{r, save_eden_raster}
writeRaster(eden_raster,file.path(".","processed","eden.tif"),overwrite=TRUE)
writeRaster(urban_raster,file.path(".","processed","urban.tif"),overwrite=TRUE)
```

## Digital Elevation Model

In the plot of the loaded data we can see that the DEM extends beyond the boundary of the
catchment. The processing in `dynatopGIS` treats all the finite DEM values as
being within the catchment. This means we need to trim the DEM to the
catchment boundaries.

To achieve this we will mask the DEM with the rasterised catchment, so
setting all values not in the catchment to `NA`.

```{r, mask_dem}
masked_dem <- mask(dem,eden_raster)
plot(masked_dem)
plot(eden,add=TRUE)
```

Save the masked DEM for later use:

```{r, save_eden_dem}
writeRaster(masked_dem,file.path(".","processed","dem.tif"),overwrite=TRUE)
```

> To those of familiar with the raster package might have noticed that `edenDEM` could have created in one line by
> ```{r, mask_dem_2,eval=FALSE}
> edenDEM <- rasterize(eden, dem, mask=TRUE)
> ```

## Channel Network

The representation of the channel network in dynatop is as a series of
connected reach lengths. In the processing done by dynatopGIS Each reach is represented by a single spatial line or polygon. The data for each reach must include an identifier for the
head and foot of the reach to allow the connectivity to be derived as well as
the reach length. Examining the data for the example with

```{r channel_data}
head(channel@data)
```

we see that 

- an identifier for the foot of the reach is provided by endNode
- an identifier for the head of the reach is provided by startNode
- the reach length is given in provided by length

Within the data there is also a unique identifier which can be used to relate
the gauges to the channel network.

## Locating Gauges on the Channel Network

Before locating the gauges on the network note that in the map of the loaded data there is a single gauge outside to the catchment
area. As simple example of manipulating spatial data in R is to remove this by
cropping the gauges to the catchment area. This is done by

```{r crop_gauges}
gauges <- crop(gauges,eden)
```

To determine the location of the gauges on the river network we identify for each gauge the
closest reach. The identifier of this reach, along with the distance to
the gauge data is then added to the gauge data.

```{r locate_gauges}
gauges$chn_identifier <- character(nrow(gauges)) # initialise storage of the channel identifier
gauges$chn_distance <- numeric(nrow(gauges)) # initialise storage of the distance to the channel

## loop to compute nearest reach for each gauge
for(ii in 1:nrow(gauges)){
  tmp <- rgeos::gDistance(gauges[ii,], channel, byid=TRUE)
  jj <- which.min(tmp)
  gauges$chn_identifier[ii] <- channel$identifier[jj]
  gauges$chn_distance[ii] <- tmp[jj]
} 
```

To check the results by plotting:

```{r crop_gauges_plot}
selected <- channel[channel$identifier %in% gauges$chn_identifier,]
plot(eden) # outlines of subcatchments in black
plot(channel,add=TRUE,col="blue") # channel network in blue
plot(selected,add=TRUE,col="orange") # selected reachs in orange
plot(gauges,add=TRUE,col="red",pch=21) # gauges as red filled circles
```

The cropped list of gauges with channel identifier can be saved to a shapefile:

```{r save_cropped_gauges}
shapefile(gauges,file.path(".","processed","gauges"))
```

> The quality of results of any method used for any automatic method of
> locating the gauges depends upon the accuracy of the GIS data. 

> In the
> numeric solution used in dyantop it is assumed that a gauge is at the foot of
> a reach. This should be allowed from when conceptualising the channel reaches.
