---
title: "Processing Observed data"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
knitr::opts_knit$set(root.dir = "./eden_data")
```

## Aim

To explore and pre-process the observed data for the Eden catchment.

## Installing required R packages

Alongside the packages used in the [initial GIS
processing](./10_initial_gis.html), two further packages are required. The first,
xts, allows for an efficent representation of time series data and can be
installed with
```{r xts_install, eval=FALSE}
install.packages("xts")
```

The second extra package required is dynatop. This can be installed with

```{r dynatop_install, eval=FALSE}
install.packages("dynatop", repos = c("https://waternumbers.github.io/drat", "http://cran.rstudio.com"))
```

Once installed the required packages are attached to the R environment so that there
functions are available.
```{r attach_obs}
library(raster)
library(xts)
library(dynatop)
```
## Expected Data

The data used in dynatop when simulating the model is expected
to be a xts object (a special kind numeric matrix whose row names give the
time) with a constant time step $\Delta t$. Each column of data should have a
unique name which is used to specify the time series when building the model.

Three variables are used as inputs, precipitation, potential evapotranspiration and discharge.
It is expected that the precipitation and potential evapotranspiration inputs series are
given in $m$ accrued over the proceeding time step. So the data value given
at time $t+\Delta t$ is the total accrued in the interval between time $t$
and time $t+\Delta t$.

The data series of discharge are used for the diffuse and point inputs which
can be added to the channel network. These series should be in $m^3/s$. In the channel HRU solution the value given at time
$t+\Delta t$ is presumed to be constant over the interval $t$ to $t+\Delta t$
so ideally the the input series should represent the average flow for this period.
 
## Data Files

The example contains two data files Eden_Flow.csv and Eden_Precip.nc. The
first is a comma seperated text file of flow data, the later a netcdf file
containing a time series of spatial rainfall fields.

## Flow Data

Looking at the start of Eden_Flow.csv

```{r show_eden_flow}
cat(readLines(file.path(".","unprocessed","Eden_Flow.csv"),5),sep="\n")
```

we see that the Index column gives the time in an unabiguous format recognised by xts. The
second column the observed discharges at the Sheepmount gauge as numeric values. We read this
in using function avaialbe through the xts package

```{r read_eden_flow}
flow <- as.xts(read.zoo(file.path(".","unprocessed","Eden_Flow.csv"),header=TRUE,sep=",",drop=FALSE))
head(flow)
```

## Rainfall Data

To explore the rainfall data open the file using the raster package

```{r open_precip}
## read in as a multilayered object
precip_brk <- brick(file.path(".","unprocessed","Eden_Precip.nc"))
## show summary
precip_brk
## show start of the z values which are the time stamps as character strings
head(getZ(precip_brk))
```

From the properties of precip_brk we see that:

- there are 1392 raster layers covering February 2020
- the projection and resolution of the rainfall grids are different to the DEM
- the time step is different to that of the flow data

To address the projection and resolution differences two approaches could be
taken

- reproject the rainfall data to match the DEM, so giving each DEM
cell its own rainfall time series
- assign each cell in the DEM to one of the rainfall data raster cells and
reformat the exisiting rainfall data to the give time series.

Since the DEM has a higher spatial resolution then the rainfall the
second approach is more computationally efficent and is recommended.

To impliment the approach start by creating a raster layer of the same
projection and extent as the rainfall feilds but with unique values

```{r rainfall_id}
rid <- raster(precip_brk[[1]])
rid <- setValues(rid,1:ncell(rid))
rid
```

Then extract the precipitation values to give an xts object where each column is the time
series for one value in rid.

```{r rainfall_out}
precip <- getValues(precip_brk) # matrix where each row is one cell
## add row names based on unique values in rid to give time series names
rownames(precip) <-  paste0("precip_",getValues(rid))
## get times as characters and convert to R internal time representation (POSIXct)
precip_datetime <- as.POSIXct(getZ(precip_brk),tz="GMT")
## create the xts object
precip <- xts(t(precip),order.by=precip_datetime)
```

To fix the difference in time step compared with the flow data we resample the precipitation data. The resample_xts function of dynatop provides
a simple method for doing this

```{r resample_time}
## call the resample_precip function
resampled_precip <- resample_xts(precip,900)
## look at what it does (assigned rainfall equally between new steps in same time period)
head(precip[,"precip_48"]) # original data
head(resampled_precip) # resampled data - see how they sum to the original values
```

To assign the DEM cells to rainfall series reproject the raster layer of unique
rainfall series numbers (rid) to match the projection and resolution of the DEM
using nearest neighbours, then save this for later use.

```{r resample_space}
## load the dem
dem <- raster(file.path(".","processed","dem.tif"))
## call the resample_precip function
reproj_rid <- projectRaster(rid,dem,method="ngb")
## mask so only values for finite DEM cells
reproj_rid <- mask(reproj_rid,dem)
## A plot for visualisation
plot(reproj_rid)
## save
writeRaster(reproj_rid,file.path(".","processed","precip_id.tif"))
```

## Potential Evapotranspriation

If available spatial Potential Evapotranspiration time series can be handled
in exactly teh same way as the rainfall. However if there is only a single series generatation of the map of
id values can be avoided. A simple sinusoidal PET series can be generated
using the evap_est function in dynatop
```{r pet}
## use a maximum of 3 mm/day - output in meters
pet <- evap_est(index(flow), eMin = 0, eMax = 0.003)
names(pet) <- "PET" # ensure series has a name
```

## Combining Series

The xts package has a convient functions for merging xts objects by the
time. We will save this as an R object for later use

```{r merge}
obs <- merge(flow,resampled_precip,pet,all=FALSE) ## merge by time stamp
saveRDS(obs,file.path(".","processed","obs.rds"))
```

> **WARNING** 
> dynatop does not allow for missing data. Any missing values must
> be replaced during the pre-processing
