---
title: "Model Building"
---

```{r, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
knitr::opts_knit$set(root.dir = "./eden_data")
```

## Aim

To process the GIS data into a Dynamic TOPMODEL structure of the Eden catchment.

## Required R packages

This exercise builds upon the early [initial GIS](./10_initial_gis.html) and
[input](./11_observed_series.html) processing, using the outputs generated during those
exercises. 

The packages used can be attached with

```{r attach_packages}
library("dynatopGIS")
library("terra")
```

and remember to move to the eden_data directory e.g.

```{r, eval=FALSE}
setwd("../eden_data")
```

## Getting started

The `dynatopGIS` packages works through a number of steps to generate a dynamic
TOPMODEL object suitable for use in with the `dynatop` package. Each step
generates one or more layers which are saved as raster or shape files into the projects
directory (which is not necessarily the R working directory).

Initialise the analysis by creating a new `dynatopGIS` object specifying the location
of the project directory, which will be created if it doesn't exist.

```{r, initialisation}
ctch <- dynatopGIS$new(file.path(".","dynaGIS"))
```

If existing files are present in the directory this step will try to reload
them as a dynatopGIS project.

## Adding catchment data

The basis of the analysis is a raster catchment map. Onto this a Digital Elevation Model (DEM) of
the catchment and a vector representation of the river network with
attributes will be added. Currently these can be in any format supported by the `terra`package.

However, within the calculations used for sink filling,
flow routing and topographic index calculations the DEM is presumed
to be projected so that is has square cells such that
the difference between the cell centres (in meters) does not alter.

The catchment map needs to be loaded first

```{r add_catchment}
ctch$add_catchment( file.path(".","processed","eden.tif") )
```

Either the DEM or channel files can be added to the project next. In this
case add the DEM processed during the initial GIS analysis:

```{r, add_dem}
ctch$add_dem(file.path(".","processed","dem.tif"))
```

As part of loading the DEM any blocks of `NA` values within the DEM that fall
within the catchment are filled to appear as sinks. These will be filled in a
later step.

Adding the channel data is achieved with
```{r channel_data}
ctch$add_channel(file.path(".","processed","channel.shp"))
```

## Getting and plotting catchment information

The `dynatopGIS` class has methods for returning and plotting the GIS data in
the project. The names and file locations of all the different GIS layers stored by dynatopGIS is returned by

```{r, list_layers}
ctch$get_layer()
```

Note that channel_vect layer is the vector representation of the channel,
while channel layer is a rasterised version.

Individual layers can be plotted (with or without the channel), for example

```{r, plot}
ctch$plot_layer("dem", add_channel=TRUE)
```

or returned to the R workspace, for example

```{r, get_layer}
ctch$get_layer("dem")
```

## Filling sinks

For the hill slope to be connected to the river network all DEM cells must
drain to those that intersect with the river network. The algorithm
implemented in the `sink_fill` method ensures this is the case. 
In calling the `sink_fill` method a flow direction algorithm
is specified and the resulting flow paths recorded. If sub-catchments are
present in the catchment map then only flow paths within the sub-catchment are
considered.

Since the algorithm used by the `sink_fill` method is iterative and the execution time of the
function is limited by capping the maximum number of iterations. If
this limit is reached without completion the method can call again with the
"hot start" option to continue from where it finished.

For the Eden the DEM is already well conditions and the algorithm
only alters a small areas of the catchment.

```{r, sink_fill}
ctch$sink_fill()

terra::plot( ctch$get_layer('filled_dem') - ctch$get_layer('dem'),
             main="Changes to height")
```

## Determining Ordering

The computational scheme in the `dynatop` package works with an ordered
sequence of HRUs constructed such that the sequence moves downslope to
catchment outlet. This is achieved by banding the channel reaches and
hillslope cells such that the catchment outlet(s) are in band 1, those cells
or reaches draining only into band 1 are in band 2 and so forth. Banding is
achieved by the following call

```{r, band}
ctch$compute_band()
ctch$plot_layer("band")
```

## Computing Properties

Two sets of properties are required for Dynamic TOPMODEL. The first set is
those required within the evaluation of the Hill slope HRU; gradient and contour
length. The second set are those used for dividing the catchment up into
different response classes. Traditionally the summary used for the
separation of the classes is the topographic index, which is the natural
logarithm of the up slope area divided by gradient.

These properties are computed using the formulae in [Quinn et al. 1991](https://doi.org/10.1002/hyp.3360050106).

The upstream area is computed by routing down slope with the fraction of the
area being routed to the next downstream pixel being proportional to the
gradient times the contour length.

The local value of the gradient is computed using the average of a
subset of between pixel gradients. For a normal 'hill slope' cell these are the
gradients to down-slope pixels weighted by contour length.

These properties are computed in an algorithm that passes over
the data once in descending height. It is called as follows

```{r, calc_atb}
ctch$compute_properties()
```

The plot of the topographic index shows a pattern of increasing values closer
to the river channels

```{r, plot_atb}
##plot of topographic index (log(a/tan b))
ctch$plot_layer('atb')
```

Although not used in ordering the HRUs `dynatopGIS` also provides the ability to compute flow distances for the hill slope cells.
The calculation of three distances is supported

- *Shortest flow length* - the shortest length based on the pixel flow paths
to a channel
- *Dominant flow length* - the distance to a channel moving in the dominant
(largest fraction) flow direction from any grid cell
- *Expected flow length* - the distance to the channel based on a weighted
average of the down-slope flow lengths. Weights are given by the fraction of
flow in each direction.

The computation, in this example for the shortest flow length, is initiated with
```{r, flow_length}
ctch$compute_flow_lengths(flow_routing="shortest")
```

The additional layers can be examined as expected
```{r, flow_length_plot}
ctch$get_layer()
ctch$plot_layer("shortest_flow_length")
```

## Adding additional Layers

Properties for use in the classification of HRUs or for determining input
series may be added as additional raster GIS layers. 

In the initial GIS analysis a layer for urban area was developed, while processing the input data produced a map relating to rainfall input.
These can now be added to the dynatopGIS project by specifying the file
location and name to call the layer.

```{r, add_class_layer}
ctch$add_layer(file.path(".","processed","urban.tif"),"urban")
ctch$add_layer(file.path(".","processed","precip_id.tif"),"precip_id")
```

Calling the get_layer method shows these are added to the project

```{r get_layer_add_layer}
ctch$get_layer()
```


## Classifying into Hydrological Response Units

Methods are provided for the classification of the catchment. 
Multiple classifications can be produced but that used to generate the
`dynatop` model (see following sections) must include the "band" layer.
 
By definition each channel length is treated as a single class with its own
Channel HRU. Two ways are
provided for dividing the hill-slope up into classes.
The first way is *cutting* where a landscape property is
divided up into classes. 
The second is *combining* where unique existing class layers are combined by
either by pairing (where unique combinations of classes form a new class) or
burning (which enforces classes onto distinct areas).

### Cutting

To split a catchment into HRUs using cuts the breaks between the classes need to be
specified. The breaks can be specified as either:

- A single value: defines the number of splits which are automatically selected
- A vector of values: treated as the breaks between classes

To demonstrate the use of cuts alone dividing the topographic index into 20
classes. Specifying a single value gives class breaks that are equally spaced
across the range of the variable.

```{r, atb_band_cut_simple}
## simple equally spaced cuts
ctch$classify("atb_20","atb",20)
```

It is perhaps better to consider 20 classes with approximately equal land
area. it this case the breaks can be computed using the quantiles of the variable

```{r, atb_20_cut}
## use the quantile function of the raster package to compute breaks
## note we need to include the min and max
brk <- global(ctch$get_layer("atb"), fun=quantile,probs = seq(0, 1, length = 21),na.rm=T)
ctch$classify("atb_20_equal","atb",brk)
```

In the meta data the cuts and burns used to generate the layer are recorded. 
These can be retrieved with

```{r see_class}
head(ctch$get_method("atb_20"))
## returns the list of break points for the classes
```

### Combining

The sub-catchments could be added alongside the banding to add a further spatial separation.

```{r, atb_20_sub}
ctch$combine_classes("atb_20_subcatch",c("atb_20","catchment"))
```

If this is to be used to generate a model the `band` layer would have to be included.

```{r, atb_20_sub_band}
ctch$combine_classes("atb_20_subcatch_band",c("atb_20_subcatch","band"))
```

A more complex combination would also include the urban area.
Layers burnt into the classification are added after the cuts are applied and
treated as directly defined HRUs. In the example it may be desirable to burn
in the Urban areas since these will have a different hydrological behaviour.

Since the layers burnt in are applied with no alteration to their values these
may clash with those already defined by the cuts. One method
that *nearly* always avoid this is to burn in classes with negative
values. Let us then adapt the urban layer to have negative values.

```{r neg_urban}
tmp <- -ctch$get_layer("urban")
writeRaster(tmp,file.path(".","processed","neg_extended_eden.tif"),overwrite=TRUE) ## write out
ctch$add_layer(file.path(".","processed","neg_extended_eden.tif"),"neg_urban")
```

then burn this into the existing classification based on the topographic index
and sub-catchment

```{r, atb_band_cut_urban_burn}
ctch$combine_classes("atb_20_subcatch_band_burn_neg_urban","atb_20_subcatch_band",burns="neg_urban")
```

In the meta data the classes of the original class values in the combination are stored
These can be retrieved with

```{r see_class_comb}
tmp <- ctch$get_method("atb_20_subcatch")
tmp$type
head(tmp$groups)
## returns the list of break points for the cuts and layers burnt in
```

### Complexity

The number of classes resulting from the classifications reflects the
complexity of the model. The number of classes can be computed from the
maps and is shown in the following table. Since these classes may occur in any
of the bands the resulting number of HRUs, which is also shown is much higher.


```{r num_hru, echo=FALSE, results='asis'}
ctch$combine_classes("atb_20_band",c("atb_20","band")) ## so we can complete the table
ctch$combine_classes("atb_20_subcatch_burn_neg_urban","atb_20_subcatch",burn="neg_urban") ## so we can complete the table

fnc <- function(nm){ length(unique(ctch$get_layer(nm))[[1]] ) }
    
tmp <- data.frame(classification = c("atb_20","atb_20_subcatch","atb_20_subcatch_band_burn_neg_urban"),
                  Classes = c( fnc("atb_20"), fnc("atb_20_subcatch"), fnc("atb_20_subcatch_burn_neg_urban")),
                  HRU = c( fnc("atb_20_band"), fnc("atb_20_subcatch_band"), fnc("atb_20_subcatch_band_burn_neg_urban"))
                  )
knitr::kable(tmp)
```


## Generating a dynamic TOPMODEL

A Dynamic TOPMODEL suitable for use with the `dynatop` package can be
generated using the create_model method. This uses an existing
classification to generate the model. 

For example, in the case of the division of by topographic index into
21 classes and the bands directly the resulting model can be generated by

```{r, model_atb_split}
ctch$create_model(file.path("./dyna","atb_20_model"), # name of new model
                  "atb_20_band", # classification to base model on
                  sf_opt="cnst",
                  sz_opt="exp",
                  rain_layer = "precip_id", # layer of input precipitation series ID
                  rain_label = "precip_" # characters added to values in rain_layer to get series name
                  )
```

Looking at the files within the folder containing the dynatopGIS project

```{r, model files}
list.files(file.path("./dyna"),pattern="atb_20_model")
```

shows that an additional raster map of the HRUs has been created in
`atb_20_model.tif` along with a file `atb_20_model.rds` containing a model
suitable for `dynatop`

## Looking at the `dynatopGIS` model output

The model output by `dynatopGIS` should be adequate for `dyantop` to run, and
is more fully documented
[here](https://waternumbers.github.io/dynatop/articles/The_Model_Object.html). Looking
at the output in more detail shows `atb_20_model.rds` contains three variables.

```{r model_1}
## load the model
mdl <- readRDS( file.path(".","dyna","atb_20_model.rds")) ## read the model in
names(mdl)
```

The `map` variable contains the path of the HRU map.
```{r map}
mdl$map
```

The `hru` variable contains a description of the HRUs. This is given as an list, with each list element representing the properties of one HRU.
```{r hru}
names(mdl$hru[[8]])
```
A summary of properties is given below, detailed descriptions are found in the [`dynatop` vignettes](https://waternumbers.github.io/dynatop/articles/The_Model_Object.html)

| Name              | Class   | Description                                              |
|-------------------|---------|----------------------------------------------------------|
| id                | integer | The unique ID of the HRU                                 |
| states            | numeric | Named numeric vector of states                           |
| properties        | numeric | Named numeric vector of properties                       |
| sf                | list    | Description of the surface zone solution                 |
| rz                | list    | Description of the surface zone solution                 |
| uz                | list    | Description of the unsaturated zone solution             |
| sz                | list    | Description of the saturated solution                    |
| sf_flow_direction | list    | Description of the surface zone outflows to other HRUs   |
| sz_flow_direction | list    | Description of the saturated zone outflows to other HRUs |
| initialisation    | numeric | Named numeric vector of initialisation parameters        |
| precip            | list    | Description of the precipitation input to the HRU        |
| pet               | list    | Description of the PET input to the HRU                  |
| class             | list    | Classification information from dynatopGIS               |


The `output_flux` variable contains a table in the format used in a call to a `dynatop` simulation to set the time series output.to set the 
```{r output_flux}
mdl$output_flux
```
By default `dynatopGIS` populates this to return the lateral outflows of the surface zones (flux column set to `q_sf`) of the catchment outlets (HRU number in the `id` column) giving then representative names.

This can be replaced by the locations of the river gauges found in the initial GIS processing.

```{r alter_model_2}
## To replace this read in the gauge locations generated mealier
gauges <- vect(file.path(".","processed","gauges.shp"))
## create a vector of the uid's of the HRUs
hid <- sapply(mdl$hru,function(h){h$id})
## create a vector of the channel names with as storaged in the class variabel
cname <- sapply(mdl$hru,function(h){h$class$name})
idx <- match(gauges$chn_identi,cname) ## index of HRU's corresponding to gauges
## create as a data.frame
gauge_output <- data.frame(name = gauges$Site.Name,
                           id = as.integer(hid[idx]),
                           flux = "q_sf",
                           scale = 1)
saveRDS(gauge_output,file.path(".","dyna","gauge_output_defn.rds")) ## save the output for later use
gauge_output

```

