---
title: "Model Building"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
knitr::opts_knit$set(root.dir = "./eden_data")
```

## Aim

To process the GIS data into a Dynamic TOPMODEL structure of the Eden catchment and
to show how to adapt this.

## Installing required R packages

This exercise builds upon the early [initial GIS
processing](./10_initial_gis.html), using the outputs generated during that
exercise. As well as the packages already used you will require the dynatopGIS
package. This can be installed with

```{r dynatopGIS_install, eval=FALSE}
install.packages("dynatopGIS", repos = c("https://waternumbers.github.io/drat", "http://cran.rstudio.com"))
```

The jsonlite package is also required and can be installed with

```{r jsonlite_install, eval=FALSE}
install.packages("jsonlite")
```


## Getting started

The dynatopGIS packages works through a number of steps to generate a dynamic
TOPMODEL object suitable for use in with the dynatop package. Each step
generates one or more layers which are saved as raster or shape files into the projects
working directory (which is not necessarily the R working directory). A record
of these layers is kept in the json format meta data file.

To start first load the library

```{r load_library}
library("dynatopGIS")
library("raster")
```

Initialise the analysis by creating a new dynatopGIS object specifying the location
of the meta data file, which will be created if it doesn't exist.

```{r, initialisation}
ctch <- dynatopGIS$new(file.path(".","dynaGIS","meta.json"))
```

## Adding catchment data

The basis of the analysis is a raster Digital Elevation Model (DEM) of
the catchment and a vector representation of the river network with
attributes. Currently these can be in any format supported by the raster
and sp packages respectively.

However, within the calculations used for sink filling,
flow routing and topographic index calculations the DEM is presumed
to be projected so that is has square cells such that
the difference between the cell centres (in meters) does not alter.

Either the DEM or channel files can be added to the project first. In this
case add the DEM processed during the initial GIS analysis:

```{r, add_dem}
ctch$add_dem(file.path(".","processed","dem.tif"))
```

As part of loading the DEM any blocks of `NA` values within the DEM are filled to
appear as sinks. This can be turned off which may significantly speed up the
loading of large DEM's. 

> **Note** An additional row and column of `NA` values is added to each edge of the
> DEM. All raster layers in the project have the same projection,
> resolution and extent of the extended DEM.

Adding river channel data is more complex. The add_channel method requires a R
object of class SpatialLinesDataFrame or SpatialPolygonsDataFrame (see the sp
package for details of these classes) which represent either spatial lines or polygons
with attribute data. Each line or polygon results in a single Channel HRU. The
data.frame of attributes must contain the following properties

- a label for the downstream end of the river length
- a label for the upstream end of the river length
- the length in meters
	
The names of the columns in the data.frame containing the required properties
are specified through the property_names input to the add_channel
method.

If the channel data is specified using spatial lines an optional extra
property; the width of the channel; can be provided. If this is not provided and is required the
value of the default_width input to the add_channel
method is used.

> **NOTE** All further columns of the data.frame are kept without change by the
> add_channel method with the exception of columns names id. This are copied to
> a column named original_id with a warning since the name id is used internally.
	
Looking at the data from the example
```{r channel_data}
channel <- shapefile(file.path(".","unprocessed","Eden_River_Network"))
head(channel@data)
```

we see that all the required properties are present, but no width is given. Using a constant width of 5 meters for all
channels can be achieved by setting

```{r add_channel_data}
ctch$add_channel(channel,property_names = c(length = "length", startNode = "startNode", endNode = "endNode"),default_width=5)
```

Alternatively an additional column containg the width could be added to the
channel data.
```{r add_channel_data_not_run, eval=FALSE}
channel$width <- 5 ## add column of widths to data frame, could be variable values
## add channels with a constant width of 5m
ctch$add_channel(channel)
## this is short for
## ctch$add_channel(channel,property_names = c(length = "length", startNode = "startNode", endNode = "endNode",width="width"))
## since the default value of property_names are valid
```

## Computing Basic Properties

So far, the DEM and channel data exist in isolation. Next, we compute some basic
summaries for each cell in the DEM, specifically:
-   land area - the area in the DEM cell covered by land
-   channel area - the area in the DEM cell covered by channel
-   channel id - the id of the channel within the cell, corresponding to
the id in the channel object.

If multiple river lengths intersect a DEM cell the properties of the channel
length with the largest area of intersection are used.

This computation is done by calling

```{r basic_properties}
ctch$compute_areas()
```

## Getting and plotting catchment information

The dynatopGIS class has methods for returning and plotting the GIS data in
the project. The names of all the different GIS layers stored by dynatopGIS is returned by

```{r, list_layers}
ctch$get_layer()
```

The complete meta data which includes the paths to the files for each layer can be retrieved with

```{r, get_meta}
head(ctch$get_meta())
```

Individual layers can be plotted (with or without the channel), for example

```{r, plot}
ctch$plot_layer("dem", add_channel=TRUE)
```

or returned to the R workspace, for example

```{r, get_layer}
ctch$get_layer("dem")
```

## Filling sinks

For the hill slope to be connected to the river network all DEM cells must
drain to those that intersect with the river network.

The algorithm implemented in the sink_fill method ensures
this is the case. Since the algorithm is iterative the execution time of the
function is limited by capping the maximum number of iterations. If
this limit is reached without completion the method can call again with the
"hot start" option to continue from where it finished.

For the Eden the DEM is already well conditions and the algorithm
only alters a small areas of the catchment.

```{r, sink_fill}
ctch$sink_fill()

raster::plot( ctch$get_layer('filled_dem') - ctch$get_layer('dem'),
             main="Changes to height")
```

## Computing properties

Two sets of properties are required for Dynamic TOPMODEL. The first set is
those required within the evaluation of the Hill slope HRU; gradient and contour
length. The second set are those used for dividing the catchment up into
different response classes. Traditionally the summary used for the
separation of the classes is the topographic index, which is the natural
logarithm of the up slope area divided by gradient.

These properties are computed using the formulae in [Quinn et al. 1991](https://doi.org/10.1002/hyp.3360050106).

The upstream area is computed by routing down slope with the fraction of the
area being routed to the next downstream pixel being proportional to the
gradient times the contour length.

The local value of the gradient is computed using the average of a
subset of between pixel gradients. For a normal 'hill slope' cell these are the
gradients to down-slope pixels weighted by contour length. In the case of
pixels which contain river channels the average of the gradients from up-slope
pixels weighted by contour length us used.

These properties are computed in an algorithm that passes over
the data once in descending height. It is called as follows

```{r, calc_atb}
ctch$compute_properties()
```

The plot of the topographic index shows a pattern of increasing values closer
to the river channels

```{r, plot_atb}
##plot of topographic index (log(a/tan b))
ctch$plot_layer('atb')
```

## Adding additional Layers

Properties for use in the classification of HRUs or for determining input
series may be added as additional
raster GIS layers. 

In the initial GIS analysis two layers where generated, one of
sub-catchments and one of urban areas. To add these we first need to extend
them to match the extent of the DEM which is padded with NA on import by dynatopGIS:

```{r, pad_class_layer}
## read in urban areas raster
tmp <- raster(file.path(".","processed","urban.tif"))
tmp <- extend(tmp,1,NA) ## extent with a row/column of NA on each edge
writeRaster(tmp,file.path(".","processed","extended_urban.tif")) ## write out

## similar for the sub-catchments
tmp <- raster(file.path(".","processed","eden.tif"))
tmp <- extend(tmp,1,NA) ## extent with a row/column of NA on each edge
writeRaster(tmp,file.path(".","processed","extended_eden.tif")) ## write out
```

These can now be added to the dynatopGIS catchment representation

```{r, add_class_layer}
ctch$add_layer("urban",file.path(".","processed","extended_urban.tif"))
ctch$add_layer("subcatch",file.path(".","processed","extended_eden.tif"))
```

A similar process allows the grid of rainfall series identifiers to be added
```{r, add_rain_layer}
tmp <- raster(file.path(".","processed","precip_id.tif"))
tmp <- extend(tmp,1,NA) ## extent with a row/column of NA on each edge
writeRaster(tmp,file.path(".","processed","extended_precip_id.tif")) ## write out
ctch$add_layer("precip_id",file.path(".","processed","extended_precip_id.tif")) ## add to dynatopGIS catchment
```

Calling the get_layer method shows these are added to the project

```{r get_layer_add_layer}
ctch$get_layer()
```

# Flow distances and ordering

Since dynatop simulations make use of ordered HRUs, a
metric is required to perform the ordering. The calculation of four such metrics is supported

- *shortest flow length* - the shortest length based on the pixel flow paths
to a channel
- *Dominant flow length* - the distance to a channel moving in the dominant
(largest fraction) flow direction from any grid cell
- *Expected flow length* - the distance to the channel based on a weighted
average of the down-slope flow lengths. Weights are given by the fraction of
flow in each direction.
- *Band* - A strict computational order, starting at the channel and arranged such
that if all pixels in band $i+1,i+2,\ldots$ are evaluated the inflows to band
$i$ are known.

The computation is initiated with
```{r, flow_length}
ctch$compute_flow_lengths()
```

The additional layers can be examined as expected -->
```{r, flow_length_plot}
ctch$get_layer()
ctch$plot_layer("band")
```

## Classifying into Hydrological Response Units

Methods are provided for the classification of the catchment. 
The classifications generated in this process are
 augmented with a further distance based separation when generating a
 dynatop model (see following section).
 
By definition each channel length is treated as a single class with its own
Channel HRU. Two ways are
provided for dividing the hill-slope up into classes.
The first way is *cutting* where one or more landscape properties are
divided up into classes and HRUs defined by each unique combination of
classes.
The second *burning* enforces classes onto distinct areas.

The classify method of a dynatopGIS allows the use of both methods, first
*cutting* then *burning* to create hill-slope HRUs.

### Cutting

To split a catchment into HRUs using cuts the breaks between the classes need to be
specified. This is done by forming a named list of cuts. The names
correspond to layers of GIS data within the dynatopGIS object. The values of
the variables should be numeric and are used as follows:

- `NA`: use the layer 'as is'; i.e. presume it is already classified
- Single value: defines the number of splits which are automatically selected
- Vector of values: treated as the breaks between classes

To demonstrate the use of cuts alone consider a model where the HRUs are
defined by the topographic index value divided into 21 classes and the band (see distance
ordering) providing a spatial ordering.

```{r, atb_band_cut}
ctch$classify("atb_band_cut",cuts=list(atb=20,band=NA))
```

The sub-catchments could be added alongside this as a further cut
```{r, atb_subcatch_band_cut}
ctch$classify("atb_subcatch_band_cut",cuts=list(atb=20,subcatch=NA,band=NA))
```

### Burning

Layers burnt into the classification are added after the cuts are applied and
treated as directly defined HRUs. In the example it may be desirable to burn
in the Urban areas since these will have a different hydrological behaviour.

Since the layer burnt in are applied with no alteration to their values the
values specified may clash with those already defined by the cuts. One method
that *nearly* always avoid this is to burn in classes with negative
values. Let us then adapt the urban layer to have negative values

```{r neg_urban}
tmp <- -ctch$get_layer("urban")
writeRaster(tmp,file.path(".","processed","neg_extended_eden.tif")) ## write out
ctch$add_layer("neg_urban",file.path(".","processed","neg_extended_eden.tif"))
```
```{r, atb_band_cut_urban_burn}
ctch$classify("atb_band_cut_neg_urban_burn",cuts=list(atb=20,band=NA),burns="neg_urban")
```

In the meta data the cuts and burns used to generate the layer are recorded. 
These can be retrieved with

```{r see_class}
ctch$get_class_method("atb_band_cut_neg_urban_burn")
## returns the list of break points for the cuts and layers burnt in
```

### Complexity

The number of Hill-slope classes resulting from the classifications reflects the
complexity of the model. The number of Hill-slope classes can be computed from the
maps

```{r num_hru, echo=FALSE, results='asis'}
tmp <- data.frame(classification = c("atb_band_cut","atb_subcatch_band_cut","atb_band_cut_neg_urban_burn"),
                  Number_HRU = rep(NA,3))
for(ii in 1:3){
  tmp$Number_HRU[ii] <- length(unique(ctch$get_layer(tmp$classification[ii])))
}
knitr::kable(tmp)
```


## Generating a dynamic TOPMODEL

A Dynamic TOPMODEL suitable for use with the dynatop package can be
generated using the create_model method. This uses an existing
classification to generate the model. 

Since dynatop simulations make use of ordered HRUs to work down-slope, a
classification which used a distance layer (see earlier section) which
represents the ordered down-slope sequencing of the pixels is recommended. 

> *It is strongly recommended that the 'band'
> distance metric is used directly as shown below*.

Even if a distance layer is not used in the classification one must be given to the create_model
method, so the resulting HRUs can be ordered.

For example, in the case of the division of by topographic index into
21 classes and the bands directly the resulting model can be generated by

```{r, model_atb_split}
ctch$create_model("atb_band_model", # name of new model
                  "atb_band_cut", # classification to base model on
                  "band", # distance to use for determining flow routing
                  transmissivity = "bounded_exponential", #transmissivity profile to use
                  rain_layer = "precip_id", # layer of input precipitation series ID
                  rain_label = "precip_" # characters added to values in rain_layer to get series name
                  )
```

Looking at the files within the folder containing the dynatopGIS project

```{r, model files}
list.files(file.path(".","dynaGIS"),pattern="atb_band_model*")
```

shows that an additional raster map of the HRUs has been created in
`atb_band_model.tif` along with a file `atb_band_model.rds` containing a model
suitable for `dynatop`

The map of HRUs can be plotted as with any layer

```{r plot_model}
ctch$plot_layer("atb_band_model")
```

The values on the map correspond to the ìd column of the hill-slope table in
the `dynatop` model.


## Model Structure

The model for use in dynatop consists of multiple data.frames. A complete
definition is given in a [dynatop
vignette](https://waternumbers.github.io/dynatop/articles/The_Model_Object.html)

The model output by dynatopGIS may not exactly match the representation of the
catchment require. One issue with the model just generated is that the PET
input is not correctly specified:

```{r alter_model_1}
## load the model
mdl <- readRDS(file.path(".","dynaGIS","atb_band_model.rds")) ## read the model in
head(mdl$pet_input) ## shows that name of the PET input series is unknown
mdl$pet_input$name <- "PET" ## when actually it was called PET in the input data
```

A further improvement is to add the gauges to the correct locations

```{r alter_model_2}
## By default gauge table contains the list of the channel HRUs that have no downstream connect
mdl$gauge 
## To replace this read in the gauge locations generated mealier
gauges <- shapefile(file.path(".","processed","gauges"))
## create a new table to replace the one in the model
new_gauge <- data.frame(name = gauges@data$Site_Nm,
                        chn_dnt = gauges@data$chn_dnt)
new_gauge$id <- as.integer( sapply(new_gauge$chn_dnt,
                                   function(x){mdl$channel$id[mdl$channel$identifier==x]}))
new_gauge$chn_dnt <- NULL ## removes the column
## add the new table to the model
mdl$gauge <- new_gauge
## then save for later use
saveRDS(mdl,file.path(".","processed","atb_band_model.rds"))

```
