---
title: "Running a Simulation"
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
knitr::opts_knit$set(root.dir = "./eden_data")
```

## Aim

To simulate a Dynamic TOPMODEL of the Eden catchment, visualise the output and
show how the parameters may be altered.

## Installing required R packages

No further packages are required beyond those used in processing the [GIS
processing](./10_initial_gis.html) and [observed](./11_observed_series.html) data.

## A Simulation

The minimum data need to perform a simulation with dynatop is a model, such
as that output by dynatopGIS in the [previous
section](./13_model_building.html) and observed data in the format introduced
[earlier](./11_observed_series.html).

Load the previously created data

```{r load_data}
mdl <- readRDS(file.path(".","processed","atb_band_model.rds"))
obs <- readRDS(file.path(".","processed","obs.rds"))
```

Attach the dynatop package so it is available in the R session then
initialise a dynatop object by loading the model

```{r load_model}
library(dynatop)
library(xts)
ctch_mdl <- dynatop$new(mdl)
```
This step performs some basis checks on the model for consistency.
The data can then be added; again simple checks are performed.

```{r add_data}
ctch_mdl$add_data(obs)
```

There are two types of HRU; hill-slope and
channel. These can be run individually with the sim_hillslope and
sim_channel methods or sequentially with the sim method. The individual
methods check that suitable input data is available, but not how it was generated.

The initial states of the simulations can be specified in the model
object. If, as in the case of this example, the states are not specified then
any attempt to perform a simulation will fail.

```{r sim_fail, error=TRUE, purl=FALSE}
## expect a failure
ctch_mdl$sim()
```

The states need to be initialised using the initialise method. This
presumes:

- The surface is dry
- The root zone is partially full (the s_rz0 parameter)
- There is a specified (see below) uniform maximum recharge rate representing the flow
  from the unsaturated to saturated zone
- the saturated zone is at a steady state
- the unsaturated zone storage ensures that the recharge rate from the
unsaturated zone to the saturated zone is as close to the maximum as possible

The uniform maximum recharge rate is input to the method call. A simple way
of getting an initial estimate of this is to divide the observed discharge at
the outlet of the catchment for the first time step by the hill-slope surface area. This works best if
the first time step has no rainfall and is close to base-flow.

```{r q0}
## determining the initial recharge rate
q0 <- obs$Sheepmount_obs[1] / sum(mdl$hillslope$area)
```

In the following we initialise the states and then perform the simulation, using the chaining of commands.

```{r initial_sim}
ctch_mdl$initialise(q0)$sim() #plot_state("s_sz") -->
```

### Mass balance

It is possible to output the mass balance for each time step of the
simulation using the `get_mass_errors` method. The returns an xts object giving
the volumes in the hill-slope states at the start and end of the time step along with the
other fluxes as volumes. This can easily be used to plot the errors as shown below.

```{r mass_check}
mb <- ctch_mdl$get_mass_errors()
head(mb) ## note the negative signs existing or outward flow to allow mass balance to be computed by summation
mb$err <- rowSums(mb) ## the total error in m^3
plot( mb$err , main="Mass balance error",ylab="m^3")
```

## Visualisations

After a simulation it is possible to get or plot output. Three types of output
are available:

- the flow at the gauges after simulating the channel HRUs
- the inflows to the channel HRUs after simulating the Hill-slope HRUs
- the current states of the system after simulating the Hill-slope Grus

All have methods for plotting the data or returning in as a variable as shown
in the following

### Gauge flows
```{r gauge_flow}
## For simulated flows at the gauge sites
sim_gauge <- ctch_mdl$get_gauge_flow() ## extract the flow as an xts object
head(sim_gauge)

## Plot the simulated flows at the gauges
ctch_mdl$plot_gauge_flow()
```

The observed gauge data is readily added and plotted
```{r gauge_flow_merge}
plot( merge(obs$Sheepmount_obs,sim_gauge),main="Simulated and observed flows",
     legend.loc='topright')
```
### Channel Inflow

```{r channel_flow}
## For inflows to the Channel HRU from the Hill-slope HRUs

channel_inflow <- ctch_mdl$get_channel_inflow() ## extract the flow as an xts object
## there is one column for each Channel HRU. The column names are the Channel HRU IDs (as strings)
## For example to see the start of the inflows to the 23,560 and 1602 channel HRUS
head(channel_inflow[,c("23","560","1602")])

## Plot the total inflow to all channel HRUs from the Hill-slopes
ctch_mdl$plot_channel_inflow(total=TRUE)
## setting total to FALSE plots all the channel HRUs inflows individually
```

While the total channel HRU inflow is useful it may also be desirable to see
if the flux is coming from the surface or saturated zones. 

```{r sep_channel_flow}
## To get the inflows to the Channel HRU from the Hill-slope HRUs separated by surface and saturated zone contributions
sep_channel_inflow <- ctch_mdl$get_channel_inflow(separate=TRUE)
## for example the start of the inflow to the Channel HRUs with ID 23, 560 and 1602
head(sep_channel_inflow$saturated[,c("23","560","1602")]) ## inflow from the saturated zone
head(sep_channel_inflow$surface[,c("23","560","1602")]) ## inflow from the surface zone

## For an overall picture
ctch_mdl$plot_channel_inflow(total=TRUE,separate=TRUE)
```

### Current States

The current states of the model can be extracted and plotted.

```{r current_states}
## extract the current states as a data.frame
current_state <- ctch_mdl$get_states()
## columns correspond to the Hill-slope HRU ID, and the storage's in each of the four zones
head(current_state)

## a single state can be plotted at one time
ctch_mdl$plot_state("s_sz")
```

## Changing Parameters

The parameters are stored in the model as which is made up of standard R
data.frame's. 
Using the properties within the data frame simple parameter changes are
straightforward

```{r simple_parameter_change}
## Some simple manipulations of the Hill-slope HRU parameters

## set all Hill-slope HRU exponential decay (m) parameters to 0.03
mdl$hillslope$m <- 0.03

## set all Hill-slope HRU maximum depth parameters (D) to 10m
## where the average topographic index (atb_bar) is > 20
mdl$hillslope$D[ mdl$hillslope$atb_bar > 20 ] <- 10
```

To change a parameter for all the HRUs with a specific value of one of the
classification variables requires more work

```{r complex_parameter_change}
## A more complex change - set the surface velocity (c_sf) for all HRUs in band 1 to 1 m/s

## First read in the band and HRU id maps
band <- raster(file.path(".","dynaGIS","band.tif"))
hru_id<- raster(file.path(".","dynaGIS","atb_band_model.tif"))
## find the unique HRU id values which have a band value of 1
u_hru_id <- unique(hru_id[band==1]) ## find unique hru id value
## in case we want to use the information again lets add an extra column to the model hillslope table
mdl$hillslope$in_band_1 <- FALSE ## be careful with the name in case it classes with an existing property, parameter or state
mdl$hillslope$in_band_1[mdl$hillslope$id %in% u_hru_id] <- TRUE
## Now change the parameter
mdl$hillslope$c_sf[mdl$hillslope$in_band_1] <- 1
## to see the changes
head(mdl$hillslope,20)
```

## Extracting Intermediate States

Unless requested dynatop keeps only the current state during a
simulation. To request that states are other time steps are kept a vector of times
must be passed in the call to the sim or sim_hillslope method

```{r keep_states}
## Let us repeat the same simulation
## keeping the states for every other time step
keep_times <- index(obs)[seq(2,nrow(obs),by=2)] ## an vector of time steps
ctch_mdl$initialise(q0)$sim(keep_states=keep_times) ## passing the vector of time steps to the call
```

The record of the states can know be extracted. While there is an entry for
every time step only those request will be populated

```{r state_record}
## extract the record of the states at each time step
state_rec <- ctch_mdl$get_states(record=TRUE) ## returns the whole record of states
head(names(state_rec)) ## the elements in state_rec are named after the time
head(state_rec[[format(keep_times[1])]]) ## those in keep_times are populated
head(state_rec[[format(index(obs)[1])]]) ## those not in keep_times are empty
```

The visualisation of the state record is possible either as a time series

```{r time_series_states}
## Create a time series plot of the saturated zone deficit
s_sz_rec <- lapply(state_rec,function(x){x$s_sz}) ## extra the saturated deficit from each time step
s_sz_rec <- do.call(rbind,s_sz_rec) ## convert them into a matrix
head(rownames(s_sz_rec)) ## the row names are the the times carried over from the names in state_rec
s_sz_rec <- as.xts(s_sz_rec) ## convert to and xts object using the rownames as times
names(s_sz_rec) <- paste(state_rec[[format(keep_times[1])]]$id) ## give column names as the HRU id
## for a the Hill-slope HRUs with IDs 2487, 3598 and 5201 the start of the time series are
head(s_sz_rec[,c("2487","3598","5201")])
## and a plot can be generated
plot(s_sz_rec[,c("2487","3598","5201")],main="Saturated zone deficit for select HRU",legend.loc="topright")
```

or as a map

```{r time_series_states_map}
## Create a map of the root zone storage at 2020-02-03 02:00:00
s_rz_val <- state_rec[["2020-02-03 02:00:00"]][,c("id","s_rz")] ## extract the id and rz storage
## create a map of values by substituting s_rz for the id in the hru_id map
s_rz_map<- raster::subs(hru_id, s_rz_val) ## substitute values into the map of HRU id numbers
## and plot
plot(s_rz_map,main="Root zone storage at 2020-02-03 02:00:00")
```

## Hot Starting and Saving

Hot starting; that is starting a simulation with the final states of a
previous simulation can be achieved by simply not re-initialising the
dynatop object before the next call to sim

```{r sim2}
## initialise and simulate the model
ctch_mdl$initialise(q0)$sim()
head(ctch_mdl$get_states()) ## state values after the initial run
## run again without initialising
## This will use the final states of the last simulation as the starting states
ctch_mdl$sim()
head(ctch_mdl$get_states()) ## different state values after the second run
```

If you want to save the model and current states at the end of a simulation do
not save the dynatop object. Instead extract and save he model including the
current states and save that.

```{r get_model}
new_mdl <- ctch_mdl$get_model() ## get the model structure from the dynatop object
head(new_mdl$hillslope) ## as the original model but will final states included
```


