---
title: "Running a Simulation"
---
  
```{r, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
knitr::opts_knit$set(root.dir = "./eden_data")
```

## Aim

To simulate a Dynamic TOPMODEL of the Eden catchment, visualise the output showing how the 
show how the parameters may be altered.

## Required R packages

This exercise builds upon the early [model creation](./13_model_building.html), using the outputs generated during that
exercise.

The packages used can be attached with

```{r attach_packages}
library("dynatop")
library("terra")
library("xts")
```

and remember to move to the eden_data directory e.g.

```{r, eval=FALSE}
setwd("./eden_data")
```

## A Simulation

The minimum data need to perform a simulation with dynatop is a model, such
as that output by `dynatopGIS` in the [previous
section](./13_model_building.html) and observed data in the format introduced
[earlier](./11_observed_series.html).

Load the previously created data

```{r load_data}
mdl <- readRDS(file.path(".","dyna","atb_20_model.rds"))
output_defn <- readRDS(file.path(".","dyna","gauge_output_defn.rds"))
obs <- readRDS(file.path(".","processed","obs.rds"))
```

A dynatop object is initialised by loading the model

```{r load_model}
hru <- mdl$hru
ctch_mdl <- dynatop$new(hru)
```
This step performs some basic checks on the model for consistency.
The data can then be added; again simple checks are performed.

```{r add_data}
ctch_mdl$add_data(obs)
```

The initial states of the HRUs can be specified in the model
object. If, as in the case of this example, the states are not specified then
any attempt to perform a simulation will fail.

```{r sim_fail, error=TRUE, purl=FALSE}
## expect a failure
ctch_mdl$sim(output_defn)
```

The states need to be initialised using the initialise method. This
presumes:

- The surface is dry
- The root zone is partially full (the s_rz0 parameter)
- There is a specified (see below) recharge rate representing the flow
  from the unsaturated to saturated zone
- the saturated zone is at a steady state
- the unsaturated zone storage ensures that the recharge rate from the
unsaturated zone to the saturated zone is as close to the maximum as possible

The maximum recharge rate and fraction of root zone storage full for each HRU forms part of its definition.

```{r init_defn}
hru[[1000]]$initialisation
## s_rz_0 - fraction of root zone full
## r_us_sz_0 - recharge unsaturated to saturated zone in m/s and
```

A simple way of getting an initial estimate of the recharge is to divide the observed discharge at
the outlet of the catchment for the first time step by the catchment area. This works best if
the first time step has no rainfall and is close to base-flow.

```{r q0}
## determining the initial recharge rate ad setting it in the model
r0 <- as.numeric(obs$Sheepmount_obs[1]) / sum( sapply( hru, function(x){x$properties["area"]} ) )
hru <- lapply(hru,function(h){h$initialisation["r_uz_sz_0"] <- r0;h})
```

In the following we initialise the states and then perform the simulation, using the chaining of commands.

```{r initial_sim}
## the map input is only needed for visualisation
ctch_mdl <- dynatop$new(hru,map=mdl$map)$add_data(obs)$initialise()$sim(output_defn)
```

### Mass balance

It is possible to output the mass balance for each time step of the
simulation using the `get_mass_errors` method. The returns an `xts` object giving
the volumes in the hill-slope states at the start and end of the time step along with the
other fluxes as volumes. This can easily be used to plot the errors as shown below.

```{r mass_check}
mb <- ctch_mdl$get_mass_errors()
head(mb)
plot( mb$err , main="Mass balance error",ylab="m^3")
```

## Visualisations

After a simulation it is possible to get or plot output. Two types of output
are available:

- the time series variable specified in the `output_defn` used in the simulation
- the states of the system at end of the simulation

Both have methods for plotting the data or returning in as a variable as shown
in the following

### Gauge flows
```{r gauge_flow}
## For simulated flows at the gauge sites
sim_gauge <- ctch_mdl$get_output() ## extract the flow as an xts object
head(sim_gauge)

## Plot the simulated flows at the gauges
ctch_mdl$plot_output()
```

The observed gauge data is readily added and plotted
```{r gauge_flow_merge}
plot( merge(obs$Sheepmount_obs,sim_gauge),main="Simulated and observed flows",
     legend.loc='topright')
```

### Current States

The current states of the model can be extracted and plotted.

```{r current_states}
## extract the current states as a data.frame
current_state <- ctch_mdl$get_states()
## columns correspond to the Hill-slope HRU ID, and the storage's in each of the four zones
head(current_state)

## a single state can be plotted at one time
ctch_mdl$plot_state("s_sz")
```

## Changing Parameters

The parameters are stored within the HRU descriptions; particularly the `sf`, `rz`, `uz` and `sz` elements. Each of these contains a `type` variable and corresponding named vector of parameters.
The various types and parameters for each zone are described in the (`dynatop` vignettes)[https://waternumbers.github.io/dynatop/articles/HRU.html#surface-zone-representations]. 

The most straight forward way for altering the parameters is two write a function that acts on each HRU, altering the parameters with reference to the properties of the HRU. 
The following presents a simple function of setting a faster surface velocity and kinematic solution for the river channels.

```{r simple_parameter_change}
fpar <- function(h){ ## h is the hru
    ## test if a channel
    if( !is.na(h$class$startNode) ){
        ## change type
        h$sf$type <- "kin"
        h$sf$param["c_sf"] <- 1
    }
    h
}
```


## Channel Inflow

To use `dyantop` to drive a hydraulic model the inflow from the subsurface and hillslope HRUs to the surface zone of each channel HRU may need to be computed.
The first term is (directly available )[https://waternumbers.github.io/dynatop/articles/The_Model_Object.html#output_flux], though will the wrong sign and as a volume.
Start by working out the id of each channel HRU


```{r channel_id}
chn_id <- do.call(c, lapply( hru, function(h){ if(!is.na(h$class$startNode)){ h$id } }))
```

Next start creating an output definition for the flux from root zone

```{r hyd_part1}
out_hyd <-  data.frame( name = paste0("hyd_",chn_id),
                       id = chn_id,
                       flux = "v_sf_rz", scale =-1/900) # scale is negative since flux usually downwards, 1/900 is 1/timestep
```

Next we will construct the inflow from HRUs representing hillslopes and add these to the data frame above.

```{r hyd_part2}
## function to compute the relevent part so the output definition for each potential source HRU
fout <- function(h){
    ## test if it is a channel HRU - in which case return NULL
    if( !is.na(h$class$startNode) ){ return(NULL) }
        
    ## index of flow directiuons going to other channels
    idx <- h$sf_flow_direction$id %in% chn_id
    if(!any(idx)){ return(NULL) } ## quit if no links!
    ## create partial output defn
    data.frame(
        name = paste0("hyd_",h$sf_flow_direction$id[idx]), ## which inflow does it alter
        id = h$id, ## flux comes from this HRU
        flux = "q_sf", # surface flux out of this HRU is input to the next
        scale = h$sf_flow_direction$fraction[idx] # scale is fraction of flow (since taking it away)
    )
}
## row bind to give a data.frame
tmp <- do.call( rbind, lapply( hru, fout) )

## row bind onto the intial data frame
out_hyd <- rbind( out_hyd, tmp )

## sort this, not needed but easier to look at
out_hyd <- out_hyd[ order(out_hyd$name), ]
```

In the resulting output definition multiple fluxes contribute to the same output series, combined by a summation weighted by the scale value.
Simulation and visualisation work as before.

```{r inflow_sim}
ctch_mdl <- dynatop$new(hru)$add_data(obs)$initialise()$sim(out_hyd)
plot( ctch_mdl$get_output()[,1:10])
```

## Extracting Intermediate States

Unless requested `dynatop` keeps only the final state during a
simulation. To request that states are other time steps are kept a vector of times
must be passed in the call to the `sim` method.

```{r keep_states}
## Let us repeat the same simulation
## keeping the states for every other time step
keep_times <- index(obs)[seq(2,nrow(obs),by=96)] ## an vector of time steps
ctch_mdl$initialise()$sim(output_defn,keep_states=keep_times) ## passing the vector of time steps to the call
```

The record of the states can know be extracted. While there is an entry for
every time step only those request will be populated

```{r state_record}
## extract the record of the states at each time step
state_rec <- ctch_mdl$get_states(record=TRUE) ## returns the whole record of states
head(names(state_rec)) ## the elements in state_rec are named after the time
head(sapply(state_rec,length)) ## those in keep_times are populated the remainder are empty
```

The visualisation of the state record is possible either as a time series

```{r time_series_states}
## Create a time series plot of the saturated zone deficit
s_sz_rec <- state_rec[ sapply(state_rec,length)>0 ]
s_sz_rec <- lapply(s_sz_rec, function(x){ setNames( sapply(x,function(z){z$states["s_sz"]}), sapply(x,function(z){z$id}) )} )
s_sz_rec <- do.call(rbind,s_sz_rec) ## bind to make matrix
s_sz_rec <- as.xts(s_sz_rec) ## convert to and xts object using the rownames as times
## for a the Hill-slope HRUs with IDs 2487, 3598 and 5201 the start of the time series are
head(s_sz_rec[,c("2487","3598","5201")])
## and a plot can be generated
plot(s_sz_rec[,c("2487","3598","5201")],main="Saturated zone deficit for select HRU",legend.loc="topright")
```
or as a map

```{r time_series_states_map}
## Create a map of the root zone storage at 2020-02-03 02:00:00
s_rz_val <- do.call(rbind, lapply( state_rec[["2020-02-03 00:30:00"]], function(x){ data.frame(id=x$id,s_rz = x$states["s_rz"]) }) )
## create a map of values by substituting s_rz for the id in the hru_id map
hru_id <- rast( file.path(".","dyna","atb_20_model.tif") ) ## open map from from file names stored in the model
s_rz_map<- subst(hru_id, from=s_rz_val$id, to = s_rz_val$s_rz) ## substitute values into the map of HRU id numbers
## and plot
plot(s_rz_map,main="Root zone storage at 2020-02-03 00:30:00")
```

## Hot Starting and Saving

Hot starting; that is starting a simulation with the final states of a
previous simulation can be achieved by simply not re-initialising the
dynatop object before the next call to sim

```{r sim2}
## initialise and simulate the model
ctch_mdl$initialise()$sim(output_defn)
r1 <- ctch_mdl$get_output() ## get the output from the first run
## run again without initialising
## This will use the final states of the last simulation as the starting states
ctch_mdl$sim(output_defn)
r2 <- ctch_mdl$get_output() ## get the output from the second
plot(r1$Sheepmount)
lines(r2$Sheepmount,col="red")## different state values after the second run
```

If you want to save the model and current states at the end of a simulation do
not save the dynatop object. Instead extract and save he model including the
current states and save that.

```{r get_model}
new_mdl <- ctch_mdl$get_model() ## get the model structure from the dynatop object
new_mdl[[1]] ## as the original model but will final states included
```


