---
title: "Building dynatop"
---

```{r, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
knitr::opts_knit$set(root.dir = "./eden_data")
```
## Aim

To build the dynatop package source code into a complete package

## Build Process

Due to the level of automation building the dynatop package consists of 5
lines of R code, which are generally best run in a new R session.

```{r build, eval=FALSE}
pacPath <- file.path(".","dynatop") ## path of the folder containing the package DESCRIPTION file
## the <package>::<function> format used below calls installed R packages without fully attaching then to the workspace
Rcpp::compileAttributes(pacPath) ## compile the attributes of the C++/R linkage
devtools::document(pacPath) ## compile the function documentation and NAMESPACE
devtools::check(pacPath) ## check the package by building it and running tests
## The check function produces significant amounts of output. Significant notes, warning and errors are repeated at the end
devtools::build(pacPath) ## If there are no warnings or errors in the check then call this to build the package
```

The make_dynatop.R script in the build_scripts directory contains these lines
along with other testing the build of the website using the pkgdown library and
building on other architectures using the r-hub services.
